---
title: "Opioid Mortality Risk in Georgia Counties"
author: "Feiran Zhang"
output:
  flexdashboard::flex_dashboard:
    theme: flatly
    orientation: rows
    social: menu
    source_code: "https://github.com/feiran924/opioid-dashboard"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(flexdashboard)
library(leaflet)
library(plotly)
library(dplyr)
library(sf)
library(spdep)
library(INLA)
library(RColorBrewer)
library(htmltools)
library(DT)
```

```{r load-and-model, include=FALSE, cache=TRUE}
# ── Load data ─────────────────────────────────────────────────────────────────
data_path <- file.path("data", "ga_opioid_data_updated_final.rds")
if (!file.exists(data_path)) {
  stop("Data file not found at: ", data_path,
       "\nPlace 'ga_opioid_data_updated_final.rds' in a 'data/' subfolder.")
}
ga_data <- readRDS(data_path)

# ── Prepare modeling variables ────────────────────────────────────────────────
ga_data[["Y"]]         <- ga_data[["mortality22"]]
statewide_rate         <- sum(ga_data$mortality22, na.rm=TRUE) /
                          sum(ga_data$population, na.rm=TRUE)
ga_data[["E"]]         <- ga_data$population * statewide_rate
ga_data[["county_id"]] <- seq_len(nrow(ga_data))
ga_df <- as.data.frame(ga_data)

# ── Build spatial graph ───────────────────────────────────────────────────────
ga_nb      <- poly2nb(ga_data, queen = TRUE)
graph_file <- tempfile(fileext = ".txt")
nb2INLA(graph_file, ga_nb)
ga_graph   <- inla.read.graph(graph_file)

# ── Fit full BYM2 model ───────────────────────────────────────────────────────
model_full <- inla(
  Y ~ 1 + pct_black_std + dist_to_su_std + pct_poverty_std +
    vacancy_rate_std + unemployment_rate_std + dist_to_mh_std +
    dist_to_otp_std + dist_to_treatment_std + dist_to_bupren_std +
    dist_to_hrsa_std + dist_to_usroad_std +
    f(county_id, model = "bym2", graph = ga_graph,
      hyper = list(
        phi  = list(prior = "logitbeta", param = c(1, 1)),
        prec = list(prior = "pc.prec",   param = c(1, 0.01)))),
  family            = "poisson",
  data              = ga_df,
  E                 = ga_df$E,
  control.predictor = list(compute = TRUE))

# ── Attach posterior RR ───────────────────────────────────────────────────────
ga_data[["RR"]]       <- model_full$summary.fitted.values[["mean"]]       / ga_df$E
ga_data[["RR_lower"]] <- model_full$summary.fitted.values[["0.025quant"]] / ga_df$E
ga_data[["RR_upper"]] <- model_full$summary.fitted.values[["0.975quant"]] / ga_df$E

# ── Display variables ─────────────────────────────────────────────────────────
has_poverty      <- "pct_poverty"      %in% names(ga_data)
has_unemployment <- "unemployment_rate" %in% names(ga_data)
has_dist_su      <- "dist_to_su"       %in% names(ga_data)
has_pct_black    <- "pct_black"        %in% names(ga_data)

ga_data[["pov_display"]]   <- if (has_poverty)      ga_data[["pct_poverty"]]      else ga_data[["pct_poverty_std"]]
ga_data[["unemp_display"]] <- if (has_unemployment) ga_data[["unemployment_rate"]] else ga_data[["unemployment_rate_std"]]
ga_data[["su_display"]]    <- if (has_dist_su)      ga_data[["dist_to_su"]]       else ga_data[["dist_to_su_std"]]
ga_data[["black_display"]] <- if (has_pct_black)    ga_data[["pct_black"]]        else ga_data[["pct_black_std"]]

pov_label   <- if (has_poverty)      "Poverty Rate (%)"             else "Poverty Rate (z-score)"
unemp_label <- if (has_unemployment) "Unemployment Rate (%)"        else "Unemployment Rate (z-score)"
su_label    <- if (has_dist_su)      "Dist. to SU Facility (miles)" else "Dist. to SU Facility (z-score)"
black_label <- if (has_pct_black)    "% Black Population"           else "% Black Population (z-score)"

# ── Risk category ─────────────────────────────────────────────────────────────
ga_data[["risk_cat"]] <- cut(
  ga_data$RR,
  breaks = c(0, 0.8, 1.0, 1.25, Inf),
  labels = c("Low (RR < 0.8)", "Below Average (0.8–1.0)",
             "Above Average (1.0–1.25)", "High (RR > 1.25)"),
  right  = TRUE
)

# ── Convert to WGS84 for leaflet ──────────────────────────────────────────────
ga_wgs84 <- st_transform(ga_data, crs = 4326)
```


Dashboard {data-icon="fa-map"}
=====================================

Row {data-height=80}
-------------------------------------

### About This Dashboard

<div style="font-size:14px; line-height:1.6; padding:4px;">
This dashboard visualizes opioid mortality risk across Georgia's 159 counties using a Bayesian BYM2 spatial model. **Widget 1** shows the geographic distribution of county-level relative risk. **Widget 2** explores how poverty and treatment access relate to risk. Click, hover, and filter to explore.
</div>

Row {data-height=540}
-------------------------------------

### Widget 1: Geographic Distribution of Opioid Mortality Risk {data-width=550}

```{r widget1-map}
pal <- colorNumeric(
  palette = "RdYlBu",
  domain  = c(0.4, 2.5),
  reverse = TRUE,
  na.color = "#808080"
)

tooltip_html <- sprintf(
  paste0(
    "<div style='font-family:Arial; font-size:13px; min-width:200px;'>",
    "<b style='font-size:14px;'>%s County</b><br>",
    "<hr style='margin:4px 0;'>",
    "<b>Posterior Relative Risk:</b> %.2f<br>",
    "<b>95%% Credible Interval:</b> (%.2f, %.2f)<br>",
    "<b>Observed Opioid Deaths (2022):</b> %d<br>",
    "<b>Expected Deaths:</b> %.1f<br>",
    "<b>%s:</b> %s<br>",
    "<b>%s:</b> %s<br>",
    "</div>"
  ),
  ga_wgs84$NAME,
  round(ga_wgs84$RR, 2),
  round(ga_wgs84$RR_lower, 2),
  round(ga_wgs84$RR_upper, 2),
  ga_wgs84$Y,
  round(ga_wgs84$E, 1),
  su_label,  ifelse(has_dist_su,
                    paste0(round(ga_wgs84$su_display, 1), " miles"),
                    paste0(round(ga_wgs84$su_display, 2), " (z)")),
  pov_label, ifelse(has_poverty,
                    paste0(round(ga_wgs84$pov_display, 1), "%"),
                    paste0(round(ga_wgs84$pov_display, 2), " (z)"))
)

leaflet(ga_wgs84, options = leafletOptions(minZoom = 6)) |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    fillColor    = ~pal(pmin(pmax(RR, 0.4), 2.5)),
    fillOpacity  = 0.8,
    color        = "white",
    weight       = 0.8,
    opacity      = 1,
    highlight    = highlightOptions(
      weight = 2.5, color = "#333", fillOpacity = 0.95, bringToFront = TRUE),
    label        = ~paste0(NAME, " County — RR: ", round(RR, 2)),
    labelOptions = labelOptions(style = list("font-size" = "12px"), textsize = "12px"),
    popup        = tooltip_html
  ) |>
  addLegend(
    position  = "bottomright",
    pal       = pal,
    values    = c(0.4, 2.5),
    title     = HTML("<b>Opioid Mortality<br>Relative Risk (RR)</b><br>
                     <span style='font-size:10px;font-weight:normal'>
                     RR = 1: state average<br>RR &gt; 1: higher than average<br>
                     Hover county for details</span>"),
    labFormat = labelFormat(digits = 1),
    opacity   = 0.9
  )
```

### Widget 2: Socioeconomic Context of High-Risk Counties {data-width=450}

```{r widget2-scatter}
plot_df <- as.data.frame(ga_data) |>
  select(NAME, RR, RR_lower, RR_upper, Y, E,
         pov_display, unemp_display, su_display, black_display, risk_cat) |>
  mutate(
    RR_rounded    = round(RR, 2),
    pov_rounded   = round(pov_display, if (has_poverty) 1 else 2),
    unemp_rounded = round(unemp_display, if (has_unemployment) 1 else 2),
    su_rounded    = round(su_display, if (has_dist_su) 1 else 2),
    black_rounded = round(black_display, if (has_pct_black) 1 else 2),
    tooltip = paste0(
      "<b>", NAME, " County</b><br>",
      "<b>Relative Risk:</b> ", RR_rounded,
      " (95% CrI: ", round(RR_lower,2), "–", round(RR_upper,2), ")<br>",
      "<b>Observed Deaths (2022):</b> ", Y, "<br>",
      "<b>", pov_label, ":</b> ", pov_rounded,  if (has_poverty) "%" else " (z)", "<br>",
      "<b>", unemp_label, ":</b> ", unemp_rounded, if (has_unemployment) "%" else " (z)", "<br>",
      "<b>", su_label, ":</b> ", su_rounded,     if (has_dist_su) " miles" else " (z)", "<br>",
      "<b>", black_label, ":</b> ", black_rounded, if (has_pct_black) "%" else " (z)"
    )
  )

risk_colors <- c(
  "Low (RR < 0.8)"           = "#2166ac",
  "Below Average (0.8–1.0)"  = "#74add1",
  "Above Average (1.0–1.25)" = "#f46d43",
  "High (RR > 1.25)"         = "#a50026"
)

x_title <- if (has_poverty) "Poverty Rate (%)" else "Poverty Rate (standardized z-score)"
y_title <- "Opioid Mortality Risk (RR; 1 = Georgia average)"

plot_ly(
  data      = plot_df,
  x         = ~pov_rounded,
  y         = ~RR_rounded,
  color     = ~risk_cat,
  colors    = risk_colors,
  size      = ~su_rounded,
  sizes     = c(5, 22),
  type      = "scatter",
  mode      = "markers",
  marker    = list(opacity = 0.8, line = list(width = 0.5, color = "white")),
  text      = ~tooltip,
  hoverinfo = "text"
) |>
  layout(
    title = list(
      text = paste0(
        "<b>Poverty Rate vs. Opioid Mortality Risk</b><br>",
        "<sup>Georgia Counties, 2022 — Point size = Distance to Nearest SU Facility (miles)</sup>"
      ),
      font = list(size = 14), x = 0.5
    ),
    xaxis = list(
      title     = list(text = x_title, font = list(size = 12)),
      tickfont  = list(size = 10),
      gridcolor = "#eeeeee", showgrid = TRUE, zeroline = FALSE
    ),
    yaxis = list(
      title     = list(text = y_title, font = list(size = 12)),
      tickfont  = list(size = 10),
      gridcolor = "#eeeeee", showgrid = TRUE
    ),
    shapes = list(list(
      type = "line", x0 = 0, x1 = 1, xref = "paper",
      y0 = 1, y1 = 1,
      line = list(color = "#888888", width = 1.2, dash = "dash")
    )),
    legend = list(
      title         = list(text = "<b>Risk Category</b>", font = list(size = 11)),
      font          = list(size = 10),
      itemclick     = "toggleothers",
      itemdoubleclick = "toggle"
    ),
    hoverlabel = list(bgcolor = "white", font = list(size = 11), bordercolor = "#cccccc"),
    margin = list(b = 40)
  )
```

Row {data-height=80}
-------------------------------------

### Project Impact

<div style="font-size:13px; line-height:1.6; padding:6px 10px; background:#f8f9fa; border-left:4px solid #18BC9C;">
Opioid overdose is a leading cause of preventable death in the U.S., with rural and underserved communities facing disproportionate burden due to limited treatment access. This dashboard enables public health officials to identify high-risk Georgia counties and understand how treatment facility access compounds socioeconomic disadvantage, supporting more equitable resource allocation.
</div>


About the Data {data-icon="fa-database"}
=====================================

Row
-------------------------------------

### Dataset Description

**Data Source:** County-level opioid mortality and socioeconomic data for Georgia's 159 counties, compiled from the Georgia Department of Public Health, U.S. Census Bureau American Community Survey (ACS 5-year estimates, 2018–2022), SAMHSA Treatment Locator, and HRSA data warehouse; all facility distance data are publicly accessible at [SAMHSA.gov](https://www.samhsa.gov) and [data.hrsa.gov](https://data.hrsa.gov).

**Sample Size:** N = 159 counties (complete census of all Georgia counties; no sampling).

**Data Collection:** Opioid death counts for 2022 were extracted from Georgia DPH vital statistics records; socioeconomic covariates were obtained from ACS 5-year survey estimates; treatment facility distances were computed as straight-line distances from each county centroid to the nearest SAMHSA-registered substance use, methadone, buprenorphine, mental health, and HRSA federally qualified health center facility.

**Study Population:** All Georgia residents who died from opioid-related causes in 2022, analyzed at the county level; county-level socioeconomic covariates describe the resident population of each Georgia county during the same period.

**Time Period & Geography:** Calendar year 2022; statewide coverage of Georgia, United States (all 159 counties).

---

### Model Details

The interactive visualizations display results from a **BYM2 (Besag-York-Mollié 2) spatial model** fitted via **INLA** (Integrated Nested Laplace Approximation) with a Poisson likelihood. The model includes:

- **2 focal variables:** % Black population, distance to nearest substance use treatment facility
- **9 confounders:** poverty rate, vacancy rate, unemployment rate, distance to mental health facility, distance to OTP clinic, distance to any treatment facility, distance to buprenorphine provider, distance to HRSA facility, distance to US road

Expected deaths (E) are computed using the statewide opioid mortality rate as an internal standard. Relative risk (RR) for each county is defined as the posterior mean fitted count divided by expected count; RR > 1 indicates elevated risk relative to the Georgia average.

---

### Impact Statement

> Opioid overdose remains a leading cause of preventable death in the United States, claiming over 80,000 lives annually, with rural and underserved communities facing disproportionate burden due to limited access to treatment. This dashboard enables public health officials and policymakers to identify high-risk Georgia counties and understand how treatment facility access compounds socioeconomic disadvantage, supporting more equitable and data-driven resource allocation.
